Poincaré and SimBio: a versatile and extensible Python ecosystem for modeling systems.

### 2024-03-22 article.tex
Update explanation of performance evaluation

To evaluate SimBio's reproducibility,
we used the \ac{SBML} test suite \cite{SBMLTestSuite},
which provides a set of \ac{SBML} models
and the expected result of a simulation.
Excluding models that use not yet supported \ac{SBML} feature,
every simulation returned correct results within the solver tolerances.

To evaluate SimBio's performance,
we used curated \ac{SBML} models from BioModels \cite{malik-sheriffBioModels15Years2020}.
Among the first 250 curated models,
we considered the 117 that used supported \ac{SBML} features.
We ran simulations using COPASI, Tellurium/RoadRunner, and SimBio.
Within SimBio, both NumPy and Numba
backends were considered.
The LSODA solver was used for COPASI and SimBio,
while for Tellurium the comparable CVODE solver was used.
In all cases, we used relative and absolute tolerances of \(10^{-6}\).
We measured three simulation stages:
loading, the initial (cold) run, and subsequent (warm) runs for each model.

For COPASI and Tellurium/RoadRunner,
we noted that its runtime depended on the number of evaluation points,
something that does not seem to happen with SimBio (Figure~\ref{fig-runtime}, left).
While SimBio's NumPy backend is slower than both COPASI and RoadRunner,
we obtained an order of magnitude speed-up using the numba backend
putting it on par with them.
A user might have to consider the trade-off between compilation and run times,
as the compilation of the \ac{RHS} code might take longer than the runtime itself,
and not be worth it for running the model only once.
Another speed-up in the runtime can be had by switching the LSODA \texttt{scipy} solver
for a more efficient \texttt{numbalsoda} implementation that avoids the Python interpreter
between each of the integration steps.
This last combination beats all other methods,
which is also true for the other models we tested (Figure~\ref{fig-runtime}, right).

>>>>

To evaluate SimBio's reproducibility,
we used the \ac{SBML} test suite \cite{SBMLTestSuite},
which provides a set of \ac{SBML} models
and the expected result of a simulation.
Excluding models that use \ac{SBML} features not yet supported by SimBio,
every simulation returned correct results within the solver tolerances.

To evaluate SimBio's performance,
we selected \ac{SBML} models from the curated section of BioModels \cite{malik-sheriffBioModels15Years2020}.
Among the first 250 models,
we considered the 117 that used supported \ac{SBML} features.
We ran simulations using
COPASI (with BasiCO) with the LSODA solver,
RoadRunner (with Tellurium) with the comparable CVODE solver, and
SimBio with the LSODA solver.
For SimBio, we considered three variants:
NumPy backend and scipy's LSODA solver,
Numba backend and scipy's LSODA solver,
Numba backend and numbalsoda's LSODA solver.
In all cases, we used a relative tolerance of \(10^{-9}\) and absolute tolerances of \(10^{-9}\).
We measured two simulation stages:
an initial cold run that includes the reading of the \ac{SBML} model
and subsequent warm runs.

For COPASI and Tellurium,
we noted that its runtime depended on the number of evaluation points,
something that does not seem to happen with SimBio (Figure~\ref{fig-runtime}, left).
While SimBio's NumPy backend is slower than both COPASI and RoadRunner,
we obtained an order of magnitude speed-up using the numba backend
putting it on par with them.
A user might have to consider the trade-off between compilation and run times,
as the compilation of the \ac{RHS} code might take longer than the runtime itself,
and not be worth it for running the model only once.
Another speed-up in the runtime can be had by switching the LSODA \texttt{scipy} solver
for a more efficient \texttt{numbalsoda} implementation that avoids the Python interpreter
between each of the integration steps.
This last combination beats all other methods,
which is also true for the other models we tested (Figure~\ref{fig-runtime}, right).
